abxD01 Paper

### 2015-04-10 Schubert_abxD01_mBio_2015.Rmd
Incorporate Schloss edits

Perturbations to the gut microbiota result in a loss of colonization resistance against gastrointestinal pathogens such as *Clostridium difficile*. Although *C. difficile* infection is commonly associated with antibiotic use, the precise alterations to the microbiota associated with this loss in function are unknown. We used a variety of antibiotic perturbations to generate a diverse array of gut microbiota structures, which we then challenged with *C. difficile* spores. Across these treatments we observed that *C. difficile* resistance was never attributable to a single organism, but rather it was the result of multiple microbiota members interacting in a context-dependent manner. Using relative abundance data, we built machine learning regression models to predict the levels *C. difficile* that were found 24 hours after challenging the perturbed communities. This model was able to explain `r round(rsq_full_features, 1)`% of the variation in the observed number of *C. difficile* per gram of feces. This model revealed important bacterial populations within the microbiota, which correlation analysis alone did not detect. Specifically, we observed that populations associated with the *Porphyromonadaceae*, *Lachnospiraceae*, *Lactobacillus*, and *Alistipes* were protective and populations associated with *Escherichia* and *Streptococcus* were associated with high levels of colonization. In addition, a population affiliated with *Akkermansia* indicated a strong context dependency on other members of the microbiota. Together, these results indicate that individual bacterial populations do not drive colonization resistance to *C. difficile*. Rather, resistance can be achieved with multiple diverse assemblages.

@@

The gastrointestinal tract harbors a complex community of bacteria, known as the microbiota, which plays an integral role in resisting the invasion by gut pathogens. This resistance has been shown to be crucial for protection against *C. difficile* infections (CDI), which are the leading source of hospital-acquired infections in the United States. Antibiotics are a major risk factor for acquiring CDI due to their effect on the normal structure of the indigenous gut microbiota. We showed that diverse antibiotic perturbations gave rise to altered communities that varied in their susceptibility to *C. difficile* colonization. Instead of being dependent on one specific population of bacteria, we found that multiple co-existing populations conferred resistance. By understanding the relationships between *C. difficile* and members of the microbiota it will be possible to better manage this important infection.

@@

The microbiota, or the diverse community of microorganisms living in and on the body, has an integral role in deterring pathogen colonization and infection {Vollaard, 1994 \#683}. This native protection by the microbiota from invasive pathogenic species is termed colonization resistance. It is well established that the gut bacterial microbiota is critical in the hosts’ defense against the pathogen *Clostridium difficile* {Wilson, 1985 \#1045}{Chang, 2008 \#590}{Reeves, 2011 \#118}. When this indigenous community is perturbed, this often leads to a loss of resistance. This is especially important in many hospital settings where patients are not only exposed to various types and degrees of perturbations, such as antibiotics, diet changes, and chemotherapy, but they are also exposed to *C. difficile* spores from their environment {Bauer, 2015 \#1047}. *C. difficile* infections (CDI) are the most reported hospital-acquired infection in the United States and are responsible for 14,000 deaths a year {Lessa, 2015 \#950}.

It is not completely understood how different perturbations to the gut microbiota result in a loss of colonization resistance to *C. difficile*. There is a clear need to better understand the ecology of *C. difficile* and its interactions with members of the microbiota. In mouse models of CDI, the baseline, untreated murine microbiota is completely resistant to *C. difficile* colonization. It was previously shown that C57Bl/6 mice treated with cefoperazone {Reeves, 2011 \#118}{Theriot, 2011 \#79}, tigecycline {Bassis, 2014 \#945}, clindamycin {Buffie, 2012 \#977}, or clindamycin in combination with a five antibiotic cocktail {Chang, 2008 \#590} had decreased colonization resistance. These studies suggest that a loss of *Lachnospiraceae* and *Barnesiella* and a bloom of *Lactobacilliaceae* and *Enterobacteriaceae* are responsible for the loss of colonization resistance. These results are largely supported by human association studies {Vincent, 2013 \#37}{Schubert, 2014 \#1048}. We observed significant differences between the gut microbiota of hospitalized individuals with and without *C. difficile* and non-hospitalized controls {Schubert, 2014 \#1048}. In addition, fecal microbiota transplants increase *Bacteroidetes* and decrease *Proteobacteria* levels in recipients, resulting in a successful restoration of colonization resistance in patients {Seekatz, 2014 \#1051}. Precisely how this occurs is not fully understood, but it emphasizes the importance of the gut microbiota in colonization resistance against *C. difficile*.

Because the gut microbiota is a complex community we need tools that can enable us to dissect the interactions within the community and with *C. difficile*. One approach is the use of models to identify associations between members of the microbiota and *C. difficile*. Models have been used to predict *C. difficile* {Stein, 2013 \#988}{Schubert, 2014 \#1048} and *Citrobacter* infection {Belzer, 2014 \#1053}, colon cancer {Zackular, 2013 \#472}, and psoriasis {Statnikov, 2013 \#477} based on the composition of the gut microbiota. By using models to explain the relationship between members of the gut microbiota, we hope to identify the subset of the normal murine microbiota that are responsible for colonization resistance.

The purpose of this investigation was to expand our current knowledge of the effects of various perturbations on colonization resistance against *C. difficile*. Through the administration of different antibiotic classes, doses, recovery times we altered the murine gut microbiota and challenged the communities with C. difficile spores and observed differences in colonization resistance. We then used 16S rRNA gene sequencing to identify structural changes within the microbiota that would be predictive of colonization resistance. Using these data, we built a random forest model to predict *C. difficile* colonization levels. Through this analysis, we have identified groups of related bacteria that are associated with *C. difficile* colonization resistance. This model revealed that the interactions that gave rise to colonization resistance were non-linear and context dependent. These findings show we can successfully apply modeling techniques to accurately measure the colonization resistance ability of a given microbiota.

@@

**Antibiotics differentially alter the structure of the microbiota and their colonization resistance to* C. difficile*.** We selected a panel of seven antibiotics from six classes with the goal of differentially altering the microbiota and assessing their resistance to *C. difficile* colonization (**Table 1**). On the day after the mice were challenged with *C. difficile* we enumerated the density of *C. difficile* in the animals' feces. We observed reproducibly high levels of *C. difficile* colonization in mice treated with cefoperazone, metronidazole, and streptomycin (**Figure 1**). We observed variable levels of *C. difficile* colonization in mice treated with ampicillin. Only one of six mice receiving vancomycin was colonized with *C. difficile* and none of the mice that received ciprofloxacin were colonized. We sequenced the 16S rRNA genes from the fecal communities of treated and untreated mice prior to *C. difficile* challenge to identify populations within the microbiota that conferred colonization resistance. All of the antibiotic treatments, except for the ciprofloxacin-treated mice (AMOVA, , P=`r round(max(control_sig[control_comparison=="cipro-control"]), 2)`), resulted in distinct and reproducible changes to the structure of the microbiota relative to the untreated animals (**Figures 1 and S1**; AMOVA, , P<`r ifelse(max(control_sig[control_comparison!="cipro-control"]) < 1e-3, "0.001", "NA")`). Comparisons of the microbiota between antibiotic classes indicated that their structures were significantly different from each other (AMOVA, P<`r round(max(other_sig[other_comparison!="amp-cef"]), 2)`). The community structures of mice receiving beta-lactams (i.e. cefoperazone and ampicillin) were not significantly different from each other (AMOVA, P=`r round(max(other_sig[other_comparison=="amp-cef"]), 2)`). These results indicate that perturbing the gut microbiota with antibiotics resulted in non-overlapping community structures that yielded significant variation in susceptibility to colonization when challenged with *C. difficile*.

**Reduced perturbations result in altered levels of colonization.** Based on the *C. difficile* colonization levels in our seven antibiotic treatments, we hypothesized that titrating the dose of antibiotics that the mice received would result in smaller perturbations to the microbiota and a reduced sensitivity to *C. difficile*. In addition to the previous treatments, we treated mice with lower concentrations of cefoperazone, streptomycin, and vancomycin (**Figures 2 and S2**). These antibiotics were selected because they are thought to target a broad spectrum of bacteria (i.e. cefoperazone), Gram-negative (i.e. streptomycin), and Gram-positive (i.e. vancomycin) bacteria (**Table 1**). As expected, colonization levels decreased significantly in all mice receiving titrated doses of cefoperazone (P\<0.02; **Figure 2**). Titrating the dose of cefoperazone in the animals' drinking water resulted in significant decreases in the relative abundance of an OTU associated with the genus *Escherichia* (OTU 3) and increases in the relative abundances of an OTU associated with the family *Porphyromonadaceae* (OTU 5) and an OTU associated with the genus *Pseudomonas* (OTU 65; Figure 2). The dose response for these three OTUs qualitatively followed what we had expected based on the correlation-based analysis. Reducing the dose of streptomycin significantly reduced the colonization levels (P\<0.01; Figure 2). Titrating the dose of streptomycin in the drinking water resulted in significant changes in the relative abundance of OTUs associated with the *Porphyromonadaceae* (OTUs 2, 3, 5, 9, 10, 13), *Alistipes* (OTU 11), and *Bacteroidales* (OTU 17). In addition to its anti-Gram-positive activity, Vancomycin was also selected because although the community was quite different from untreated mice, we observed high levels of colonization in only one mouse. We anticipated that lower doses might result in a community structure that would result in colonization. In fact, the 0.3 and 0.1 mg/mL doses of vancoymicin resulted in similarly high levels of colonization (P=0.96). Seven OTUs were differentially represented across the three doses of vancomycin. Surprisingly, even though the colonization levels did not significantly differ between the mice receiving 0.1 and 0.3 mg/mL of vancomycin in their drinking water, four of the OTUs that had significantly different relative abundances were only found in the lower dose. Three of these were affiliated with members of the *Porphyromonadaceae* (OTUs 2, 3, and 5) and one was affiliated with a member of the genus *Bacteroides* (OTU 1). Two OTUs affiliated with the *Akkermansia* (OTU 6) and *Lactobacillus* (OTU 8) genera increased with increasing dose and a third OTU affiliated with *Escherichia* (OTU 4) had a mixed response to the dose level. These results suggest that the context of the microbiota is important in determining the overall resistance to *C. difficile*. For example, the relationship between the *Bacteroides* (OTU 1) and *C. difficile* colonization is positive in streptomycin-treated mice and it is negative in cefoperazone-treated mice. In addition, cefoperazone and streptomycin-treated mice have high levels of *C. difficile* although the former have significantly higher levels of *Escherichia* (OTU 4), which are absent in the streptomycin-treated mice. Together, these results suggest that individual populations were not sufficient to consistently predict colonization resistance. In light of such results, resistance is likely a product of the overall composition of the community.

@@

**Reduced perturbations result in altered levels of colonization.** Based on the *C. difficile* colonization levels in our seven antibiotic treatments, we hypothesized that titrating the dose of antibiotics that the mice received would result in smaller perturbations to the microbiota and a reduced sensitivity to *C. difficile*. In addition to the previous treatments, we treated mice with lower concentrations of cefoperazone, streptomycin, and vancomycin (Figure S2). These antibiotics were selected because they are thought to target a broad spectrum of bacteria (i.e. cefoperazone), Gram-negative (i.e. streptomycin), and Gram-positive (i.e. vancomycin) bacteria. As expected, colonization levels decreased significantly in all mice receiving titrated doses of cefoperazone (P\<`r round(max(cef.test$p.value, na.rm=T), 2)`; Figure 2). Titrating the dose of cefoperazone in the animals' drinking water resulted in significant decreases in the relative abundance of an OTU associated with the genus *Escherichia* (OTU 4) and increases in the relative abundances of an OTU associated with the family *Porphyromonadaceae* (OTU 9) and an OTU associated with the genus *Pseudomonas* (OTU 65; **Figure 2**). The dose response for these three OTUs qualitatively followed what we had expected based on the correlation-based analysis. Reducing the dose of streptomycin significantly reduced the colonization levels (P\<`r round(max(strep.test$p.value, na.rm=T), 2)`; **Figure 2**). Titrating the dose of streptomycin in the drinking water resulted in significant changes in the relative abundance of OTUs associated with the *Porphyromonadaceae* (OTUs 2, 3, 5, 9, 10, 13), *Alistipes* (OTU 11), and *Bacteroidales* (OTU 17). In addition to its anti-Gram-positive activity, Vancomycin was also selected because although the community was quite different from untreated mice, we observed high levels of colonization in only one mouse. We anticipated that lower doses might result in a community structure that would result in colonization. In fact, the 0.3 and 0.1 mg/mL doses of vancoymicin resulted in similarly high levels of colonization (P=`r round(vanc.test$p.value["0.3","0.1"], 2)`)). Seven OTUs were differentially represented across the three doses of vancomycin. Surprisingly, even though the colonization levels did not significantly differ between the mice receiving 0.1 and 0.3 mg/mL of vancomycin in their drinking water, four of the OTUs that had significantly different relative abundances were only found in the lower dose. Three of these were affiliated with members of the *Porphyromonadaceae* (OTUs 2, 3, and 5) and one was affiliated with a member of the genus *Bacteroides* (OTU 1). Two OTUs affiliated with the *Akkermansia* (OTU 6) and *Lactobacillus* (OTU 8) genera increased with increasing dose and a third OTU affiliated with *Escherichia* (OTU 4) had a mixed response to the dose level. These results suggest that the context of the microbiota is important in determining the overall resistance to *C. difficile*. For example, the relationship between the *Bacteroides* (OTU 1) and *C. difficile* colonization is positive in streptomycin-treated mice and it is negative in cefoperazone-treated mice. In addition, cefoperazone and streptomycin-treated mice have high levels of *C. difficile* although the former have significantly higher levels of *Escherichia* (OTU 4), which are absent in the streptomycin-treated mice. Together, these results suggest that individual populations were not sufficient to consistently predict colonization resistance. In light of such results, resistance is likely a product of the overall composition of the community.

@@

**Allowing recovery of the microbiota restores colonization resistance.** In the experiments we have described thus far, we have allowed the gut microbiota to recover for 24 hours before challenging them with *C. difficile*. Several studies have demonstrated that perturbed communities can return to a "healthy" state in which resistance to *C. difficile* is restored {Reeves, 2011 \#118} {Theriot, 2014 \#956}. To test the effect of recovery on colonization and gain greater insights into the populations that confer colonization resistance, we allowed the microbiota of the mice that received the full metronidazole and ampicillin treatment to recover for an additional five days (Figure S3). Among the metronidazole-treated mice, those with extended recovery had a `r format(metro_med_cfu[2]/metro_med_cfu[1], sci=T)`-fold reduction in colonization (P<`r ifelse(metro_test$p.value<0.001, "0.001", "NA")`; **Figure 3**). In addition, of the `r nrow(metro_delay)` mice given the longer recovery period, `r sum(metro_delay$CFU==0)` had no detectable *C. difficile* 24 hours after challenge. We detected six OTUs that were differentially represented in the two sets of metronidazole-treated mice (Figure 3). Most notable among these was a member of the *Barnesiella* (OTU 2) and the *Escherichia* (OTU 3). The relative abundance of this *Barnesiella* OTU increased with the delay, and the relative abundance of this *Escherichia* OTU decreased. Similar to the metronidazole-treated mice, the ampicillin-treated mice that were allowed to recover an additional five days before challenge had a significant decrease in *C. difficile* colonization (P=`r round(amp_test$p.value, 2)`). As before, we observed a similar increase and decrease in relative abundances for *Barnesiella* (OTU 2) and *Escherichia* (OTU 3). However untreated, fully resistant mice harbor significantly lower levels of *Barnesiella* OTU 2. Rather, untreated mice have high levels of various *Porphyromonadaceae* OTUs (**Figure 1**). These findings further confirm the context-dependency of colonization resistance suggested by the results of our titration experiments.

@@

**Correlation analysis reveals potentially protective bacteria.** To identify bacterial taxa that could be associated with resistance or susceptibility to *C. difficile*, we measured the correlation between the relative abundance of each OTU on the day of inoculation with the level of *C. difficile* colonization 24 hours later (**Figure 4**). OTUs associated with providing resistance against *C. difficile* (N=`r sum(corr_data$sig_corrs < 0)`) outnumbered those with associated with susceptibility (N=`r sum(corr_data$sig_corrs > 0)`). Among various bacterial families the *Porphyromonadaceae* (ρ<sub>average<\sub>=`r ave_corr_family["Porphyromonadaceae"]`, N=`r N_family["Porphyromonadaceae"]` OTUs) and *Ruminococcaceae* (ρ<sub>average<\sub>=`r ave_corr_family["Ruminococcaceae"]`, N=`r N_family["Ruminococcaceae"]` OTUs) were consistently associated with low levels of *C. difficile* colonization. Two OTUs from the *Proteobacteria* had a significant association with *C. difficile* colonization. These included OTUs associated with the genera (ρ=`r round(corr_data[families=="Pseudomonadaceae","sig_corrs"], 2)`) and *Escherichia* (ρ=`r round(corr_data[families=="Enterobacteriaceae","sig_corrs"], 2)`). These associations have been suggested previously {Reeves, 2011 \#118;Schubert, 2014 \#1048;Bassis, 2014 \#945;Buffie, 2012 \#977}. In addition, by performing an OTU-based analysis we observed several bacterial families that were represented by OTUs that were associated with low and high levels of *C. difficile* colonization. For example, the Lachnospiraceae have been associated with protection against C. difficile and although we observed `r sum(corr_data[families=="Lachnospiraceae","sig_corrs"] < 0)` OTUs that were associated with low levels of *C. difficile* and `r sum(corr_data[families=="Lachnospiraceae","sig_corrs"] > 0)` OTU that was associated with high levels of *C. difficile*. In addition we observed `r nrow(corr_data[families=="Lactobacillaceae",])` *Lactobacillus* OTUs from the *Lactobacilliaceae* where `r sum(corr_data[families=="Lactobacillaceae","sig_corrs"] < 0)` were associated with low levels of *C. difficile* and `r sum(corr_data[families=="Lactobacillaceae","sig_corrs"] > 0)` was associated with high levels. The broad taxonomic representation of OTUs associated with low levels of *C. difficile* suggests that a diverse community is required to prevent the colonization *C. difficile*.

**The composition of the disturbed gut microbiota is predictive of C. difficile colonization levels.** The three sets of experiments demonstrated that in certain contexts individual OTUs could be associated with *C. difficile* colonization, but in other contexts they were not. Correlation-based analyses cannot predict these types of context dependencies because they do not take into account the non-linearity and statistical interactions between populations. This suggests that colonization is a phenotype that is driven by multiple populations that act independently and possibly in concert to resist colonization. Therefore, we used a regression-based random forest machine learning algorithm to predict the level of *C. difficile* colonization observed in the three sets of experiments based on the composition of the microbiota at the time of challenge. The model explained `r round(rsq_full_features, 1)`% of the variation in the observed *C. difficile* colonization levels (Figures 5 and S4). When we only included the top 7 OTUs based on the percent increase in the mean squared error when each OTU was removed, the resulting model explained `r round(rsq_part_features, 1)`% of the variation in the observed *C. difficile* colonization levels. Many of the OTUs that contributed the most to the quality of the fit included members of the *Porphyromonadaceae*, *Alistipes*, *Lachnospiraceae*, *Lactobacillus*, and *Escherichia* (**Figure 6**). These results further validate the observations from the correlation-based analysis. Together these results suggest that colonization resistance is likely conferred by *Porphyromonadaceae* (OTU 15, 10, and 6), *Lachnospiraceae* (OTU 25), *Lactobacillus* (OTU 23), and *Alistipes* (OTU 12). A loss in these populations, concurrently with a gain of *Escherichia* (OTU 3) or *Streptococcus* (OTU 90), can result in increased susceptibility to infection (Figure 6). As we observed in the titration experiments, the relationship between *Akkermansia* (OTU 4) and *C. difficile* colonization appears to be context dependent. There were varying abundances of *Akkermansia* in mice regardless of the level of *C. difficile* colonization.

>>>>

Perturbations to the gut microbiota result in a loss of colonization resistance against gastrointestinal pathogens such as *Clostridium difficile*. Although *C. difficile* infection is commonly associated with antibiotic use, the precise alterations to the microbiota associated with this loss in function are unknown. We used a variety of antibiotic perturbations to generate a diverse array of gut microbiota structures, which we then challenged with *C. difficile* spores. Across these treatments we observed that *C. difficile* resistance was never attributable to a single organism, but rather it was the result of multiple microbiota members interacting in a context-dependent manner. Using relative abundance data, we built machine learning regression models to predict the levels of *C. difficile* that were found 24 hours after challenging the perturbed communities. This model was able to explain `r round(rsq_full_features, 1)`% of the variation in the observed number of *C. difficile* per gram of feces. This model revealed important bacterial populations within the microbiota, which correlation analysis alone did not detect. Specifically, we observed that populations associated with the *Porphyromonadaceae*, *Lachnospiraceae*, *Lactobacillus*, and *Alistipes* were protective and populations associated with *Escherichia* and *Streptococcus* were associated with high levels of colonization. In addition, a population affiliated with *Akkermansia* indicated a strong context dependency on other members of the microbiota. Together, these results indicate that individual bacterial populations do not drive colonization resistance to *C. difficile*. Rather, resistance and sensitivity can be achieved with multiple diverse assemblages.

@@

The gastrointestinal tract harbors a complex community of bacteria, known as the microbiota, which plays an integral role preventing its colonization by gut pathogens. This resistance has been shown to be crucial for protection against *C. difficile* infections (CDI), which are the leading source of hospital-acquired infections in the United States. Antibiotics are a major risk factor for acquiring CDI due to their effect on the normal structure of the indigenous gut microbiota. We found that diverse antibiotic perturbations gave rise to altered communities that varied in their susceptibility to *C. difficile* colonization. Instead of being dependent on one specific population of bacteria, we found that multiple co-existing populations conferred resistance. By understanding the relationships between *C. difficile* and members of the microbiota it will be possible to better manage this important infection.

@@

The microbiota, or the diverse community of microorganisms living in and on the body, has an integral role in deterring pathogen colonization and infection {Vollaard, 1994 \#683}. This native protection by the microbiota from invasive pathogenic species is termed colonization resistance. It is well established that the gut bacterial microbiota is critical in the host's defense against the pathogen *Clostridium difficile* {Wilson, 1985 \#1045; Chang, 2008 \#590;Reeves, 2011 \#118}. When this indigenous community is perturbed, this often leads to a loss of resistance. This is especially important in many hospital settings where patients are not only exposed to various types and degrees of perturbations, such as antibiotics, diet changes, and chemotherapy, but they are also exposed to *C. difficile* spores from their environment {Bauer, 2015 \#1047}. *C. difficile* infections (CDI) are the most reported hospital-acquired infection in the United States and are responsible for 14,000 deaths a year {Lessa, 2015 \#950}.

It is not completely understood how different perturbations to the gut microbiota result in a loss of colonization resistance to *C. difficile*. There is a clear need to better understand the ecology of *C. difficile* and its interactions with members of the microbiota. In mouse models of CDI, the perturbed, untreated murine microbiota is completely resistant to *C. difficile* colonization. It was previously shown that C57Bl/6 mice treated with cefoperazone {Reeves, 2011 \#118; Theriot, 2011 \#79}, tigecycline {Bassis, 2014 \#945}, clindamycin {Buffie, 2012 \#977}, or clindamycin in combination with a five antibiotic cocktail {Chang, 2008 \#590} had decreased colonization resistance. These studies suggest that a loss of *Lachnospiraceae* and *Barnesiella* and a bloom of *Lactobacilliaceae* and *Enterobacteriaceae* are responsible for the loss of colonization resistance. These results are largely supported by human association studies {Vincent, 2013 \#37; Schubert, 2014 \#1048}. We observed significant differences between the gut microbiota of hospitalized individuals with and without *C. difficile* and non-hospitalized controls {Schubert, 2014 \#1048}. In addition, fecal microbiota transplants increase *Bacteroidetes* and decrease *Proteobacteria* levels in recipients, resulting in a successful restoration of colonization resistance in patients {Seekatz, 2014 \#1051}. Precisely how this occurs is not fully understood, but it emphasizes the importance of the gut microbiota in colonization resistance against *C. difficile*.

Because the gut microbiota is a complex community we need tools that can enable us to dissect the interactions within the community and with *C. difficile*. One approach is the use of models to identify associations between members of the microbiota and *C. difficile*. Models have been used to predict *C. difficile* {Stein, 2013 \#988; Schubert, 2014 \#1048} and *Citrobacter* infection {Belzer, 2014 \#1053}, colon cancer {Zackular, 2013 \#472}, and psoriasis {Statnikov, 2013 \#477} based on the composition of the gut microbiota. By using models to explain the relationship between members of the gut microbiota, we similarly sought to identify the subset of the normal murine microbiota that are responsible for colonization resistance.

The purpose of this investigation was to expand our current knowledge of the effects of various perturbations on colonization resistance against *C. difficile*. Through the administration of different antibiotic classes, doses, and recovery times we altered the murine gut microbiota and challenged the communities with *C. difficile* spores and to quantify differences in colonization resistance. We then used 16S rRNA gene sequencing to identify structural changes within the microbiota that would be predictive of colonization resistance. Using these data, we built a random forest model to predict *C. difficile* colonization levels. Through this analysis, we have identified groups of related bacteria that are associated with *C. difficile* colonization resistance. This model revealed that the interactions that gave rise to colonization resistance were non-linear and context dependent. These findings show we can successfully apply modeling techniques to accurately measure the colonization resistance ability of a given microbiota.

@@

**Antibiotics differentially alter the structure of the microbiota and their colonization resistance to** ***C. difficile.*** We selected a panel of seven antibiotics from six classes with the goal of differentially altering the microbiota and assessing their resistance to *C. difficile* colonization (**Table 1**). On the day after mice were challenged with *C. difficile* we enumerated the density of *C. difficile* in the animals' feces. We observed reproducibly high levels of *C. difficile* colonization in mice treated with cefoperazone, metronidazole, and streptomycin (**Figures 1 and S1**). We observed variable levels of *C. difficile* colonization in mice treated with ampicillin. Only one of six mice receiving vancomycin was colonized with *C. difficile* and none of the mice that received ciprofloxacin were colonized. We sequenced the 16S rRNA genes from the fecal communities of treated and untreated mice prior to *C. difficile* challenge to identify populations within the microbiota that conferred colonization resistance. All of the antibiotic treatments, except for the ciprofloxacin-treated mice (AMOVA, , P=`r round(max(control_sig[control_comparison=="cipro-control"]), 2)`), resulted in distinct and reproducible changes to the structure of the microbiota relative to the untreated animals (AMOVA, P<`r ifelse(max(control_sig[control_comparison!="cipro-control"]) < 1e-3, "0.001", "NA")`). Comparisons of the microbiota between antibiotic classes indicated that their structures were significantly different from each other (AMOVA, P<`r round(max(other_sig[other_comparison!="amp-cef"]), 2)`). The community structures of mice receiving beta-lactams (i.e. cefoperazone and ampicillin) were not significantly different from each other (AMOVA, P=`r round(max(other_sig[other_comparison=="amp-cef"]), 2)`). These results indicate that perturbing the gut microbiota with antibiotics resulted in non-overlapping community structures that yielded significant variation in susceptibility to colonization when challenged with *C. difficile*.

@@

**Reduced perturbations result in altered levels of colonization.** Based on the *C. difficile* colonization levels in our seven antibiotic treatments, we hypothesized that titrating the dose of antibiotics that the mice received would result in smaller perturbations to the microbiota and a reduced sensitivity to *C. difficile*. In addition to the previous treatments, we treated mice with lower concentrations of cefoperazone, streptomycin, and vancomycin (Figure S2). These antibiotics were selected because they are thought to target a broad spectrum of bacteria (i.e. cefoperazone), Gram-negative (i.e. streptomycin), and Gram-positive (i.e. vancomycin) bacteria. As expected, colonization levels decreased significantly in all mice receiving titrated doses of cefoperazone (P\<`r round(max(cef.test$p.value, na.rm=T), 2)`; Figure 2). Titrating the dose of cefoperazone in the animals' drinking water resulted in significant decreases in the relative abundance of an OTU associated with the genus *Escherichia* (OTU 3) and a number of rare OTUs. We also observed increases in the relative abundances of OTUs associated with the family *Porphyromonadaceae* (OTU 5, 10, 11, 13, and 21; **Figure 2**). Reducing the dose of streptomycin significantly reduced the colonization levels (P\<`r round(max(strep.test$p.value, na.rm=T), 2)`; **Figure 2**). Titrating the dose of streptomycin in the drinking water resulted in significant changes in the relative abundance of OTUs associated with the *Porphyromonadaceae* (OTUs 2, 5, 6, 10, and 11), *Alistipes* (OTU 12), and *Bacteroidales* (OTU 17). In addition to its anti-Gram-positive activity, Vancomycin was also selected because although the community was quite different from untreated mice, we observed high levels of colonization in only one mouse. We anticipated that lower doses might result in a community structure that would result in colonization. In fact, the 0.3 and 0.1 mg/mL doses of vancoymicin resulted in similarly high levels of colonization (P=`r round(vanc.test$p.value["0.3","0.1"], 2)`)). Seven OTUs were differentially represented across the three doses of vancomycin. Surprisingly, even though the colonization levels did not significantly differ between the mice receiving 0.1 and 0.3 mg/mL of vancomycin in their drinking water, four of the OTUs that had significantly different relative abundances were only found in the lower dose. Three of these were affiliated with members of the *Porphyromonadaceae* (OTUs 2, 5, and 6) and one was affiliated with a member of the genus *Bacteroides* (OTU 1). Two OTUs affiliated with the *Akkermansia* (OTU 6) and *Lactobacillus* (OTU 8) genera increased with increasing dose and a third OTU affiliated with *Escherichia* (OTU 4) had a mixed response to the dose level. These results suggest that the context of the microbiota is important in determining the overall resistance to *C. difficile*. For example, the relationship between the *Bacteroides* (OTU 1) and *C. difficile* colonization was positive in streptomycin-treated mice and it was negative in cefoperazone-treated mice. In addition, cefoperazone and streptomycin-treated mice had high levels of *C. difficile* although the former had significantly higher levels of *Escherichia* (OTU 3), which were absent in the streptomycin-treated mice. Together, these results suggest that individual populations were not sufficient to consistently predict colonization resistance. In light of such results, resistance is likely a product of the overall composition of the community.

@@

**Allowing recovery of the microbiota restores colonization resistance.** In the experiments we have described thus far, we allowed the gut microbiota to recover for 24 hours before challenging them with *C. difficile*. Several studies have demonstrated that perturbed communities can return to a "healthy" state in which resistance to *C. difficile* is restored {Reeves, 2011 \#118; Theriot, 2014 \#956}. To test the effect of recovery on colonization and gain greater insights into the populations that confer colonization resistance, we allowed the microbiota of the mice that received the full metronidazole and ampicillin treatment to recover for an additional five days (**Figure S3**). Among the metronidazole-treated mice, those with extended recovery had a $`r format(metro_med_cfu[2]/metro_med_cfu[1], sci=T)`$-fold reduction in colonization (P<`r ifelse(metro_test$p.value<0.001, "0.001", "NA")`; **Figure 3**). In addition, `r sum(metro_delay$CFU==0)` of the `r nrow(metro_delay)` mice given the longer recovery period had no detectable *C. difficile* 24 hours after challenge. We detected six OTUs that were differentially represented in the two sets of metronidazole-treated mice (**Figure 3**). Most notable among these was a member of the *Barnesiella* (OTU 2) and the *Escherichia* (OTU 3). The relative abundance of this *Barnesiella* OTU increased with the delay, and the relative abundance of this *Escherichia* OTU decreased. Similar to the metronidazole-treated mice, the ampicillin-treated mice that were allowed to recover an additional five days before challenge had a significant decrease in *C. difficile* colonization (P=`r round(amp_test$p.value, 2)`). As before, we observed a similar increase and decrease in relative abundances for *Barnesiella* (OTU 2) and *Escherichia* (OTU 3). However untreated, fully resistant mice harbored significantly lower levels of *Barnesiella* (OTU 2). Rather, untreated mice had high levels of various *Porphyromonadaceae* OTUs (**Figure 1**). These findings further confirm the context-dependency of colonization resistance suggested by the results of our titration experiments.

@@

**Correlation analysis reveals potentially protective bacteria.** To identify bacterial taxa that could be associated with resistance or susceptibility to *C. difficile*, we measured the correlation between the relative abundance of each OTU on the day of inoculation with the level of *C. difficile* colonization 24 hours later (**Figure 4**). OTUs associated with providing resistance against *C. difficile* (N=`r sum(corr_data$sig_corrs < 0)`) outnumbered those with associated with susceptibility (N=`r sum(corr_data$sig_corrs > 0)`). The *Porphyromonadaceae* (ρ<sub>average</sub>=`r ave_corr_family["Porphyromonadaceae"]`, N=`r N_family["Porphyromonadaceae"]` OTUs) were consistently associated with low levels of *C. difficile* colonization. Among the three OTUs from the *Proteobacteria* had a significant positive association with *C. difficile* colonization, the strongest was affiliated with the *Escherichia* (OTU 3; ρ=`r round(corr_data[families=="Enterobacteriaceae","sig_corrs"], 2)`). By performing an OTU-based analysis we were able to observe intra-family and genus differences in association with *C. difficile* colonization. For example, the Lachnospiraceae have been associated with protection against *C. difficile* and although we observed `r sum(corr_data[families=="Lachnospiraceae","sig_corrs"] < 0)` OTUs that were associated with low levels of *C. difficile* and `r sum(corr_data[families=="Lachnospiraceae","sig_corrs"] > 0)` OTU that was associated with high levels of *C. difficile*. In addition we observed `r nrow(corr_data[families=="Lactobacillaceae",])` *Lactobacillus* OTUs from the *Lactobacilliaceae* where `r sum(corr_data[families=="Lactobacillaceae","sig_corrs"] < 0)` were associated with low levels of *C. difficile* and `r sum(corr_data[families=="Lactobacillaceae","sig_corrs"] > 0)` was associated with high levels. The broad taxonomic representation of OTUs associated with low levels of *C. difficile* again suggests that a diverse community is required to prevent the colonization *C. difficile*.

**The composition of the disturbed gut microbiota is predictive of** ***C. difficile*** **colonization levels.** These three sets of experiments demonstrated that in certain contexts individual OTUs could be associated with *C. difficile* colonization, but in other contexts those OTUs had the opposite or not association. This suggests that colonization is a phenotype that is driven by multiple populations that act independently and possibly in concert to resist colonization. Correlation-based analyses cannot predict these types of context dependencies because they do not take into account the non-linearity and statistical interactions between populations. Therefore, we used a regression-based random forest machine learning algorithm to predict the level of *C. difficile* colonization observed in the three sets of experiments using the composition of the microbiota at the time of challenge as predictor variables. The model explained `r round(rsq_full_features, 1)`% of the variation in the observed *C. difficile* colonization levels (Figures 5). When we only included the top `r n_part_features` OTUs based on the percent increase in the mean squared error when each OTU was removed, the resulting model explained `r round(rsq_part_features, 1)`% of the variation in the observed *C. difficile* colonization levels. The OTUs that were ranked as being the most important in defining the random forest model, further validated the observations from the correlation-based analysis (**Figures S4**). According to the random forest model, colonization resistance was associated with members of the *Porphyromonadaceae* (OTU 15, 10, 6, 18, and 11), *Lachnospiraceae* (OTU 25), *Lactobacillus* (OTU 23), *Alistipes* (OTU 12), and *Turicibacter* (OTU 9; **Figure 6**). A loss in these populations, concurrently with a gain of *Escherichia* (OTU 3) or *Streptococcus* (OTU 90), was associated with an increased susceptibility to infection (**Figure 6**). As we observed in the titration experiments, the relationship between *Akkermansia* (OTU 4) and *C. difficile* indicated that wide variation in the relative abundance of *Akkermansia* was associated with varying levels of *C. difficile*. There were varying abundances of *Akkermansia* in mice regardless of the level of *C. difficile* colonization. Finally, as indicated by the number of OTUs with relative abundances below the limit of detection, those mice could harbor varying levels of *C. difficle*. These observations bolster the hypothesis that colonization is context dependent.
